High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-02-25 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\footnotesize{A) Summary plot of individual specimen absolute paleointensity results (square symbols) and their averages and standard deviations at site level (black bars with grey uncertainty boxes) where each `AX' site is an individual anorthosite xenolith within the Beaver River diabase. All results are corrected for cooling rate with a factor of 0.74. The sites with successful experiments come from 3 regions (Carlton Peak, Silver Bay, Kennedy Creek) which would have cooled at distinct times yielding similar estimates within the each region with differences between regions. B) Regional means calculated from the specimen-level data are compared to the distribution of intensities calculated from a paleomagnetic axial dipole moment model for the past 2 million years (PADM2M; \citealp{Ziegler2011a}) at the latitude corresponding to the paleolatitude of study region (22\textdegree N). C) The mean of the 3 regional means is compared to means calculated from 3 random values drawn from the PADM2M model \cite{Ziegler2011a}. The distribution represents a total of 10,000 iterations of taking 3 random draws and calculating the mean. These comparisons highlight that these paleointensity values are strong in relation to the geomagnetic field over the past 2 million years.}}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

Given that the multiple records of a weak field cannot all be the minimum prior to inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby also explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions (Fig. \ref{fig:PINT_compilation}; \citealp{Bono2019a}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

We collected paleomagnetic cores that are 2.5 cm in diameter along the southern and eastern Beaver Bay Complex with a particular focus on acquiring paired sites of anorthosite xenoliths and their local diabase hosts during summer field seasons in 2019 and 2020. Sample cores were collected using a hand-held gasoline-powered drill and were oriented using a magnetic compass as well as a sun compass when possible. Sun compass orientations were preferentially used for determining the sample azimuth. Sister specimens underwent step-wise alternating field (AF) or thermal demagnetization at the UC Berkeley Paleomagnetism Lab to isolate paleomagnetic directions (data presented in \cite{Zhang2021b}). Based on the anorthosite thermal demagnetization results, we selected sites whose unblocking temperature ranges are narrow and near 580\textdegree C for paleointensity experiments. Beaver River diabase sites with minimal secondary remanence were also selected for paleointensity experiments.

>>>>

\begin{figure*}[h!]
\noindent\includegraphics[width=17.8 cm]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\footnotesize{A) Summary plot of individual specimen absolute paleointensity results (square symbols) and their averages and standard deviations at site level (black bars with grey 1$\sigma$ uncertainty boxes) where each `AX' site is an individual anorthosite xenolith within the Beaver River diabase. All results are corrected for cooling rate with a factor of 0.74. The sites with successful experiments come from 3 regions (Carlton Peak, Silver Bay, Kennedy Creek) which would have cooled at distinct times yielding similar estimates within the each region with differences between regions. B) Regional means calculated from the specimen-level data are compared to the distribution of intensities calculated from a paleomagnetic axial dipole moment model for the past 2 million years (PADM2M; \citealp{Ziegler2011a}) at the latitude corresponding to the paleolatitude of study region (22\textdegree N). C) The mean of the 3 regional means is compared to means calculated from 3 random values drawn from the PADM2M model \cite{Ziegler2011a}. The distribution represents a total of 10,000 iterations of taking 3 random draws and calculating the mean. These comparisons highlight that these paleointensity values are strong in relation to the geomagnetic field over the past 2 million years.}}
\label{fig:PINT_cooling_corrected}
\end{figure*}

@@

Given that the multiple records of a weak field in the Proterozoic and Paleozoic cannot all be the minimum prior to the singular event of the onset of inner core nucleation, what processes could lead to a weak dipole at Earth's surface even in the presence of a crystallizing inner core? Numerical models have shown that the dipole moment is sensitive to both the magnitude and spatial pattern of heat flow across the core-mantle boundary when there are strong available power sources to the geodynamo \cite{Olson2007a, Olson2010a}. In such models, relatively low total heat flux across the core-mantle boundary can prevent the axial dipole from reversing whereas a high heat flux through the boundary can result in an increase in reversal frequency and decrease in dipole intensity. The ``Mid-Paleozoic Dipole Low'' has been hypothesized to be the result of such elevated core-mantle boundary heat flux conditions at a time when there was also available power from a crystallizing inner core \cite{Hawkins2019a}, thereby also explaining the observed low paleointensities which include values as weak as those of the ca. 565 Ma Sept-$\hat{I}$les intrusions (Fig. \ref{fig:PINT_compilation}; \citealp{Bono2019a}). Mantle convection can modulate core mantle boundary heat flow through changes in the structure of the deep mantle associated with upwelling plumes \cite{Larson1991a, Courtillot2007a} and subducted slabs \cite{Tan2002b, Biggin2012a, Hounslow2018a}. Strong evidence for differential plate tectonic motion extends back to ca. 2.2 Ga in the Paleoproterozoic \cite{Mitchell2014a,Swanson-Hysell2021b} and potentially back to ca. 3.2 Ga in the Archean \cite{Brenner2020a}. Plate tectonic modulations of core mantle boundary heat flow are therefore expected throughout the Proterozoic. Such changes may explain large variability in Proterozoic paleointensity values similar to those seen in the Phanerozoic \cite{Lloyd2021a} and may challenge our ability to detect the increase in surface geomagnetic field strength predicted to have happened at the onset of inner core crystallization.

@@

We collected paleomagnetic cores that are 2.5 cm in diameter along the southern and eastern Beaver Bay Complex with a particular focus on acquiring paired sites of anorthosite xenoliths and their local diabase hosts during summer field seasons in 2019 and 2020. Sample cores were collected using a hand-held gasoline-powered drill and were oriented using a magnetic compass as well as a sun compass when possible. Sun compass orientations were preferentially used for determining the sample azimuth. Sister specimens underwent step-wise alternating field (AF) or thermal demagnetization at the UC Berkeley Paleomagnetism Lab to isolate paleomagnetic directions (data presented in \citealp{Zhang2021b}). Based on the anorthosite thermal demagnetization results, we selected sites whose unblocking temperature ranges are narrow and near 580\textdegree C for paleointensity experiments. Beaver River diabase sites with minimal secondary remanence were also selected for paleointensity experiments.

### 2022-02-25 PNAS_manuscript/cover_letter.tex
Updates from Overleaf

Dear \textit{Proceedings of the National Academy of Sciences} editors,

@@

In this study, we target a quite unique rock type which are lower crustal anorthosite cumulates that were brought to the near surface where they cooled and acquired their magnetization 1.1 billion years ago. In contrast to the diabase that hosts the xenoliths which exhibits non-ideal behavior, the anorthosites are high fidelity paleointensity recorders. These anorthosites are different than those typical of layered mafic intrusions as they formed deeper in the crust. We are able to use $\mu$m-scale magnetic maps generated with a quantum diamond microscope to show that in contrast to typical anorthosites with exsolved magnetite needles there is minimal grain-scale anisotropy associated with the small magnetite hosted within the plagioclase crystals. The data that emerges from this unique archive shows that the geomagnetic field was quite strong 1.1 billion years ago --- stronger than any time in the past 2 million years and rivaling some of the strongest dipole moments of the entire database. Rather than there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon, there was a strongly powered dynamo in the late Mesoproterozoic.

Multiple observed transitions from weak to strong paleointensity values in the Phanerozoic and the Proterozoic cannot all be the minimum prior to the beginning of inner core nucleation. Instead they likely indicate are the result of plate tectonics modulating core-mantle heat flow patterns in the Proterozoic as in the Phanerozoic (giving rise to intervals such as the mid-Paleozoic dipole low). We are excited about the results presented in this manuscript and believe that the readers of \textit{Proceedings of the National Academy of Sciences} will find them compelling and a significant advance.

>>>>

Dear \textit{PNAS} editors,

@@

In this study, we target a quite unique rock type which are lower crustal anorthosite cumulates that were brought to the near surface where they cooled and acquired their magnetization 1.1 billion years ago. In contrast to the diabase that hosts the xenoliths which exhibits non-ideal behavior, the anorthosites are high fidelity paleointensity recorders. These anorthosite xenoliths are different than those typical of layered mafic intrusions as they formed deeper in the crust, but then acquired their magnetization near to the surface. We are able to use $\mu$m-scale magnetic maps generated with a quantum diamond microscope to show that in contrast to typical anorthosites with exsolved magnetite needles there is minimal grain-scale anisotropy associated with the small magnetite hosted within the plagioclase crystals. The data that emerges from this unique archive shows that the geomagnetic field was quite strong 1.1 billion years ago --- stronger than any time in the past 2 million years and rivaling some of the strongest dipole moments of the entire database. Rather than there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon, there was a strongly powered dynamo in the late Mesoproterozoic.

Multiple observed transitions from weak to strong paleointensity values in the Phanerozoic and the Proterozoic cannot all be the minimum prior to the beginning of inner core nucleation. Instead, they likely are the result of plate tectonics modulating core-mantle heat flow patterns in the Proterozoic as in the Phanerozoic (giving rise to intervals such as the mid-Paleozoic dipole low). We are excited about the results presented in this manuscript and believe that the readers of \textit{PNAS} will find them compelling and a significant advance.
